### 2024-03-29 parts/correction.tex
big edits

\AP When multiple indeterminate are involved, the notion of degree must be
adapted, and to ensure that the decision procedure is effective, one must
obtain explicit bounds on the \kl{translations} that are needed. To that end,
let us now introduce the ordering over which the induction of
\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
\kl{well-quasi-ordering}, and is in particular well-founded \cite[Hoare
quasi-ordering]{SCSC12}.

@@

Let us now state, for the sake of a completeness, the equality of
$\CorrectPoly$ and $\CoveredPoly$ in the case of two indeterminates. This is
relevant because it shows that examples given in \cite{KARH77} are actually
correct, even though based on an invalid result, and it may be the case for
other works based on \cite{KARH77}.

@@

    If we fix none of the variables, then the \kl{maximal monomials}
    are \kl{non-negative} by assumption. If we fix one of the
    variables, we can assume without loss of generality that we 
    fix $X = k$ for some $k \in \Nat$.
    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
    and therefore must have a positive leading coefficient
    (which is the unique \kl{maximal monomial} in this case)
    or be constant equal to 0. In both cases, the \kl{maximal monomials}
    have positive coefficient.
    The same reasoning applies \emph{a fortiori} in the case where
    we fix the two indeterminate, leading to a constant polynomial.
\end{proof}

>>>>

\AP When multiple indeterminates are involved, the notion of degree must be
adapted, and to ensure that the decision procedure is effective, one must
obtain explicit bounds on the \kl{translations} that are needed. To that end,
let us now introduce the ordering over which the induction of
\cref{lem:correct-to-n-rat} is built. Recall that $\MaximalMonomials(P)$ is the
set of \kl{maximal monomials} of $P$, hence belongs to $\Pfin(\Monomials)$. We
order \kl{monomials} with the \kl{divisibility ordering}, making $\Monomials$ a
\kl{well-quasi-ordering} that is isomorphic to $\Nat^k$ with the product
ordering \cite[Dickson’s Lemma]{SCSC12}. We endow $\Pfin(\Monomials)$ with the
\intro{Hoare ordering}, that is, $S_1 \hoareleq S_2$ whenever for all
\kl{monomials} $M_1 \in S_1$, there exists a monomial $M_2 \in S_2$, such that
$M_1$ \kl{divides} $M_2$. The set $(\Pfin(\Monomials), \hoareleq)$ remains a
\kl{well-quasi-ordering}, and is in particular well-founded \cite[Hoare
quasi-ordering]{SCSC12}.

@@

Let us now state, for the sake of completeness, the equality of $\CorrectPoly$
and $\CoveredPoly$ in the case of two indeterminates. This is relevant because
it shows that examples given in \cite{KARH77} are actually correct, even though
based on an invalid result, and it may be the case for other works based on
\cite{KARH77}.

@@

    If we fix none of the variables, then the \kl{maximal monomials}
    are \kl{non-negative} by assumption. If we fix one of the
    variables, we can assume without loss of generality that we 
    fix $X = k$ for some $k \in \Nat$.
    Then $P(k,Y)$ is a \kl{non-negative} \emph{univariate} polynomial, 
    and therefore must have a positive leading coefficient
    (which is the unique \kl{maximal monomial} in this case)
    or be constant equal to 0. In both cases, the \kl{maximal monomials}
    have positive coefficient.
    The same reasoning applies \emph{a fortiori} in the case where
    we fix the two indeterminates, leading to a constant polynomial.
\end{proof}

### 2024-03-29 parts/intro.tex
big edits

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness:
Mealy Machines ($\intro*\Mealy$) \cite{MEAL55}, sequential functions
($\intro*\Sequential$) \cite{SCHU77}, rational functions ($\intro*\Rational$)
\cite{EILE74}, regular functions ($\intro*\Regular$) \cite{ENMA02},
and polyregular functions ($\intro*\Poly$) \cite{BOKL19}. For all of these
models, an \emph{aperiodic} counterpart can be defined (respectively,
$\intro*\AMealy$, $\intro*\ASequential$, $\intro*\ARational$,
$\intro*\ARegular$, $\intro*\SF$), lifting the correspondence between
counter-free automata, star-free languages and aperiodic monoids to the
functional setting \cite{FKT14,BOJA14,CADA15,BOKL19}. The inclusions between
these classes of functions are all known to be strict, and we depicted in
\cref{computational-models:fig} the status of the \emph{membership problem}
associated to these strict inclusions (that is, given a function $f$, decide if
$f$ can be computed by a function of a proper subclass). Note that in
\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
larger class $\mathsf{C}$. As an example, to conclude that a function can be
computed by an \emph{aperiodic sequential function} ($\ASequential$) it is
enough to prove that it is sequential ($\Sequential$), and computed by an
aperiodic polyregular function ($\SF$).

@@

\AP It was conjectured that, analogously to classes appearing in
\cref{computational-models:fig}, the following equation would hold: $\NSF =
\NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
decidability of aperiodicity for single letter output polyregular functions
since aperiodicity is decidable for $\ZSF$. We refer the reader to
\cref{previously-known-inclusions:fig} to get an overview of the different
classes of functions appearing in the paper. Arrows denote strict inclusions,
and effectiveness (both in terms of decidability and of effective
representation) is represented by thick double arrows. Inclusions that are
suspected to be effective are represented using a dashed arrow together with a
question mark.

\begin{figure}
    \centering
    \includestandalone[width=7cm]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
    }
    \label{previously-known-inclusions:fig}
\end{figure}

>>>>

\AP There have been recurrent attempts to generalize the notions of regularity
from languages to functions, leading to classes of various expressiveness, such
as\footnote{This list is far from exhaustive.} Mealy Machines ($\intro*\Mealy$)
\cite{MEAL55}, sequential functions ($\intro*\Sequential$) \cite{SCHU77},
rational functions ($\intro*\Rational$) \cite{EILE74}, regular functions
($\intro*\Regular$) \cite{ENMA02}, and polyregular functions ($\intro*\Poly$)
\cite{BOKL19}. For all of these models, an \emph{aperiodic} counterpart can be
defined (respectively, $\intro*\AMealy$, $\intro*\ASequential$,
$\intro*\ARational$, $\intro*\ARegular$, $\intro*\SF$), lifting the
correspondence between counter-free automata, star-free languages and aperiodic
monoids to the functional setting
\cite{FKT14,BOJA14,CADA15,DJR16,BDK18,BOKL19,DGK21}. The inclusions between
these classes of functions are all known to be strict, and we depicted in
\cref{computational-models:fig} the status of the \emph{membership problem}
associated to these strict inclusions (that is, given a function $f$, decide if
$f$ can be computed by a function of a proper subclass). Note that in
\cref{computational-models:fig}, if $\mathsf{B}$ is a proper subclass of
$\mathsf{C}$, then the \emph{aperiodic} variant of $\mathsf{B}$ is precisely
the intersection of $\mathsf{B}$ with the \emph{aperiodic} variant of the
larger class $\mathsf{C}$. As an example, to conclude that a function can be
computed by an \emph{aperiodic sequential function} ($\ASequential$) it is
enough to prove that it is sequential ($\Sequential$), and computed by an
aperiodic polyregular function ($\SF$).

@@

\AP It was conjectured that, analogously to classes appearing in
\cref{computational-models:fig}, the following equation would hold: $\NSF =
\NPoly \cap \ZSF$ \cite[Conjecture 7.61]{DOUE23}. This would imply the
decidability of aperiodicity for single letter output polyregular functions
since aperiodicity is decidable for $\ZSF$. We refer the reader to
\cref{previously-known-inclusions:fig} to get an overview of the different
classes of functions appearing in the paper.

\begin{figure}
    \centering
    \includestandalone[width=7cm]{tikz/class-inclusions}
    \caption{
        Decidability and inclusions of classes of functions,
        arranged along two axes. The first one is the complexity
        of the output alphabet ($\Nat$, $\Rel$, $\Sigma$). The second
        one is the allowed computational power
        (star-free polyregular functions, polyregular functions, 
        rational series).
        Arrows denote strict inclusions,
        and effectiveness (both in terms of decidability and of effective
        representation) is represented by thick double arrows. Inclusions that are
        suspected to be effective are represented using a dashed arrow together with a
        question mark.
    }
    \label{previously-known-inclusions:fig}
\end{figure}

### 2024-03-29 parts/starfree.tex
big edits

\AP
% Connection with Weighted Automata
% introduce weighted automata and rational series here
% but whether it can be leveraged remains unclear.
Recently, aperiodicity has been shown decidable over a restricted class of
\kl{polyregular functions} where the output belongs to $\Rel$ instead of
$\Nat$, that are called \kl{$\Rel$-polyregular functions} ($\intro*\ZPoly$)
\cite{CDTL23}. These functions generalize \kl{$\Nat$-polyregular functions}
by allowing negative outputs, and therefore enabling a limited form of
backtracking. As for the unary outputs, these results generalize to
\kl{polyregular functions} having outputs in the free commutative \emph{group}
generated by a finite alphabet $\Gamma$. The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). Although the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic characterization of \kl{star-free $\Rel$-polyregular
functions} ($\intro*\ZSF$) would continue to characterize \kl{star-free
$\Nat$-polyregular functions} ($\intro*\NSF$). 
% UNCLEAR

@@

>>>>

@@

\AP
The decidability of aperiodicity for
\kl{$\Rel$-polyregular functions} relies on the construction of a canonical
object called the \emph{residual transducer}, the latter being essentially
based on differences between functions, crucially leveraging negative outputs
(i.e., the presence of inverse elements). Although the proof method does not carry
from $\Rel$-output functions to $\Nat$-output functions, it was conjectured
that the semantic property of being \kl{ultimately polynomial}
would also characterize $\NSF$ inside $\NPoly$, a conjecture that
is restated hereafter.
